Well-quasi-orderings on word languages

### 2025-03-24 src/infixes.tex
minor changes + funding

\begin{remark}
    \label{dw-closure-not-wqo:rem}
    Let $(X,\preceq)$ be a quasi-ordered set, and $L \subseteq X$ be such that $(L,
    \preceq)$ is \kl{well-quasi-ordered}. It is not true in general that
    $(\dwset{L}, \preceq)$ is \kl{well-quasi-ordered}. In the case of $(\Sigma^*,
    \infleq)$ a typical example is to start from an infinite \kl{antichain} $A$,
    together with an enumeration $\seqof{w_i}$ of $A$, and build the language $L
    \defined \setof{ \prod_{i = 0}^n w_i }{ i \in \Nat }$. By definition, $L$ is a
    \kl{chain} for the \kl{infix} ordering, hence \kl{well-quasi-ordered}. However,
    $\dwset[\infleq]{L}$ contains $A$, and is therefore not
    \kl{well-quasi-ordered}. 
\end{remark}

@@

\begin{lemma}
    \label{uniformly-recurrent:lem}
    Let $w \in \Sigma^\Nat$ be a \kl{uniformly recurrent} word.
    Then, the set of finite \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{lemma}
\begin{proof}
    Let $L$ be the set of finite \kl{infixes} of $w$.
    Consider a sequence $\seqof{u_i}$ of words in $L$. Without loss of
    generality, we may consider a subsequence such that $\card{u_i} <
    \card{u_{i+1}}$ for all $i \in \Nat$. Because $\ThueMorse$ is \kl{uniformly
    recurrent}, there exists $k \geq 1$ such that $u_1$ is an \kl{infix} of
    every word $v$ of size at least $k$. In particular, $u_1$ is an \kl{infix}
    of $u_k$, hence the sequence $\seqof{u_i}$ is \kl(sequence){good}.
\end{proof}

@@

    Because the alphabet $\Sigma$ is finite, we can enumerate the words of $L$
    as $\seqof{w_i}$. Now, let us build the sequence $\seqof{u_i}$ by induction
    as follows: $u_0 = w_0$, and $u_{i+1}$ is a word that contains $u_i$ and
    $w_i$, which exists in $L$ because $L$ is \kl(subset){directed}. Since $L$ is
    \kl{well-quasi-ordered}, one can extract an infinite set of indices $I
    \subseteq \Nat$ such that $u_i \infleq u_{j}$ for all $i \leq j \in I$.

@@

    Assume that $w$ is \kl{ultimately uniformly recurrent}. Consider a sequence
    of words $\seqof{w_i}$ that are finite \kl{infixes} of $w$. Because $w$ is
    \kl{ultimately uniformly recurrent}, there exists a bound $N_0$ such that
    $w_{\geq N_0}$ is \kl{uniformly recurrent}. Let $i < N_0$, we claim that,
    without loss of generality, only finitely many words in the sequence
    $\seqof{w_i}$ can be found starting at the position $i$ in $w$. Indeed, if
    it is not the case, then we have an infinite subsequence of words that are
    all comparable for the \kl{infix relation}, and therefore a \kl{good
    sequence}, because the \kl{infix relation} is \kl{well-founded}. We can
    therefore assume that all words in the sequence $\seqof{w_i}$ are such that
    they start at a position $i \geq N_0$. But then, they are all finite
    \kl{infixes} of $w_{\geq N_0}$, which is a \kl{uniformly recurrent} word,
    whose set of finite \kl{infixes} is \kl{well-quasi-ordered}
    (\cref{uniformly-recurrent:lem}).

>>>>

\begin{remark}
    \label{dw-closure-not-wqo:rem}
    Let $(X,\preceq)$ be a quasi-ordered set, and $L \subseteq X$ be such that $(L,
    \preceq)$ is \kl{well-quasi-ordered}. It is not true in general that
    $(\dwset{L}, \preceq)$ is \kl{well-quasi-ordered}. In the case of $(\Sigma^*,
    \infleq)$ a typical example is to start from an infinite \kl{antichain} $A$,
    together with an enumeration $\seqof{w_i}[i \in \Nat]$ of $A$, and build the language $L
    \defined \setof{ \prod_{i = 0}^n w_i }{ i \in \Nat }$. By definition, $L$ is a
    \kl{chain} for the \kl{infix} ordering, hence \kl{well-quasi-ordered}. However,
    $\dwset[\infleq]{L}$ contains $A$, and is therefore not
    \kl{well-quasi-ordered}. 
\end{remark}

@@

\begin{lemma}
    \label{uniformly-recurrent:lem}
    Let $w \in \Sigma^\Nat$ be a \kl{uniformly recurrent} word.
    Then, the set of finite \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{lemma}
\begin{proof}
    Let $L$ be the set of finite \kl{infixes} of $w$.
    Consider a sequence $\seqof{u_i}[i \in \Nat]$ of words in $L$. Without loss of
    generality, we may consider a subsequence such that $\card{u_i} <
    \card{u_{i+1}}$ for all $i \in \Nat$. Because $\ThueMorse$ is \kl{uniformly
    recurrent}, there exists $k \geq 1$ such that $u_1$ is an \kl{infix} of
    every word $v$ of size at least $k$. In particular, $u_1$ is an \kl{infix}
    of $u_k$, hence the sequence $\seqof{u_i}[i \in \Nat]$ is \kl(sequence){good}.
\end{proof}

@@

    Because the alphabet $\Sigma$ is finite, we can enumerate the words of $L$
    as $\seqof{w_i}[i \in \Nat]$. Now, let us build the sequence $\seqof{u_i}[i \in \Nat]$ by induction
    as follows: $u_0 = w_0$, and $u_{i+1}$ is a word that contains $u_i$ and
    $w_i$, which exists in $L$ because $L$ is \kl(subset){directed}. Since $L$ is
    \kl{well-quasi-ordered}, one can extract an infinite set of indices $I
    \subseteq \Nat$ such that $u_i \infleq u_{j}$ for all $i \leq j \in I$.

@@

    Assume that $w$ is \kl{ultimately uniformly recurrent}. Consider a sequence
    of words $\seqof{w_i}[i \in \Nat]$ that are finite \kl{infixes} of $w$. Because $w$ is
    \kl{ultimately uniformly recurrent}, there exists a bound $N_0$ such that
    $w_{\geq N_0}$ is \kl{uniformly recurrent}. Let $i < N_0$, we claim that,
    without loss of generality, only finitely many words in the sequence
    $\seqof{w_i}[i \in \Nat]$ can be found starting at the position $i$ in $w$. Indeed, if
    it is not the case, then we have an infinite subsequence of words that are
    all comparable for the \kl{infix relation}, and therefore a \kl{good
    sequence}, because the \kl{infix relation} is \kl{well-founded}. We can
    therefore assume that all words in the sequence $\seqof{w_i}[i \in \Nat]$ are such that
    they start at a position $i \geq N_0$. But then, they are all finite
    \kl{infixes} of $w_{\geq N_0}$, which is a \kl{uniformly recurrent} word,
    whose set of finite \kl{infixes} is \kl{well-quasi-ordered}
    (\cref{uniformly-recurrent:lem}).
